Integration of advanced 3D SPECT modelling for pinhole collimators into the open-source STIR framework

### 2022-12-16 Manuscript/manuscript.tex
Added some references and cleaned up tex file

\documentclass[utf8]{FrontiersinHarvard} % for articles in journals using the Harvard Referencing Style (Author-Date), for Frontiers Reference Styles by Journal: https://zendesk.frontiersin.org/hc/en-us/articles/360017860337-Frontiers-Reference-Styles-by-Journal
%\documentclass[utf8]{FrontiersinVancouver} % for articles in journals using the Vancouver Reference Style (Numbered), for Frontiers Reference Styles by Journal: https://zendesk.frontiersin.org/hc/en-us/articles/360017860337-Frontiers-Reference-Styles-by-Journal
%\documentclass[utf8]{frontiersinFPHY_FAMS} % Vancouver Reference Style (Numbered) for articles in the journals "Frontiers in Physics" and "Frontiers in Applied Mathematics and Statistics"

%\setcitestyle{square} % for articles in the journals "Frontiers in Physics" and "Frontiers in Applied Mathematics and Statistics" 
\usepackage{url,hyperref,lineno,microtype,subcaption}
\usepackage[onehalfspacing]{setspace}
\usepackage{mhchem} % for element symbols
\usepackage{wasysym} % for diameter symbol
\usepackage{multirow}
\usepackage{threeparttable}

@@

\def\keyFont{\fontsize{8}{11}\helveticabold }
\def\firstAuthorLast{Strugari {et~al.}} %use et al only if is more than 1 author

@@

\def\Address{
$^{1}$Biomedical MRI Research Laboratory, Halifax, Canada\\
$^{2}$Department of Physics and  Atmospheric Science, Dalhousie University, Halifax, Canada\\
$^{3}$Neuroimaging Group, Barcelona$\mathrm{\beta}$eta Brain Research Center, Barcelona, Spain\\
$^{4}$Institute of Nuclear Medicine, University College London, London, UK\\
$^{5}$Department of Diagnostic Radiology, Dalhousie University, Halifax, Canada
}
% The Corresponding Author should be marked with an asterisk
% Provide the exact contact address (this time including street name and city zip code) and email of the corresponding author
\def\corrAuthor{Matthew Strugari}
\def\corrEmail{matthew.strugari@dal.ca}

\sloppy

@@

The design of the collimator in terms of hole-size, material, and overall geometry, among other factors, affects the spatial resolution and sensitivity of a SPECT system. A number of designs exist including but not limited to parallel-hole, fan-beam, converging and diverging, slit-slat, coded-aperture, and single- and multi-pinhole collimators. Therefore, the choice of collimator design is application dependent in order to channel photons of different energies, magnify or minify images, or select between imaging quality and imaging speed~\cite{nuclearfields.com}. Although parallel-hole and fan-beam collimators are conventionally used when imaging small fields-of-view (FOVs), pinhole collimators can provide a superior resolution-sensitivity tradeoff~\cite{collimators, recall Dr. Mawko's book}. Besides the successful application of pinhole-SPECT systems in small-animal imaging, there has been a resurgence in the use of pinhole collimators for clinical cardiac and brain studies and when imaging small FOVs~\cite{ozashin_clinical_2020}.

While the use of pinhole-SPECT has regained popularity in clinical and preclinical investigations of molecular imaging agents, there are no open-source software solutions available for reconstructing pinhole-SPECT datasets.  However, recent efforts have led to the integration of a 3D SPECT system matrix modelling library for pinhole collimators into the open-source Software for Tomographic Image Reconstruction (STIR). The STIR package is an object-oriented library implemented in C++ that provides a framework for research in the processing and reconstruction of emission tomography studies~\cite{thielemans_stir_2012}. Initially written for support of positron emission tomography (PET) data, STIR was previously extended to handle SPECT data with parallel- and converging-hole collimators~\cite{fuster_integration_2013a, fuster_integration_2013b}. The expansion of STIR's support for pinhole collimators marks the first open-source platform for reconstructing pinhole-SPECT datasets which is important in the advancement of molecular imaging techniques and technologies.

@@

This work aims to demonstrate the forthcoming capabilities of STIR's support for pinhole-SPECT datasets. This was achieved by integrating parts of the SPECT Reconstruction Library developed at the University of Barcelona (SRL-UB) into STIR~\cite{Refs. 2-5 in SPECTUB}. The library enables corrections for the spatially variant collimator-detector response and attenuation by incorporating their effects into the system matrix.

@@

Similar to the original \texttt{SPECTUB} implementation, the new pinhole-SPECT implementation is referred to as \texttt{PinholeSPECTUB} and includes a dedicated reader for SPECT projection data in interfile format~\cite{interfile}. The pinhole-SPECT interfile reader utilizes the projection matrix size, pixel scaling factor, and detector radius according to the face of the scintillating crystal. Calculation of the system matrix is executed with the \texttt{ProjMatrixByBinPinholeSPECTUB} projector class derived from the existing STIR \texttt{ProjMatrixByBin} class which utilizes a detector and collimator text file in addition to the usual STIR parameter file. The parameter file is a text file which uses an Interfile-like syntax. It is composed of keywords corresponding to the names of the various reconstruction and matrix parameters with the values entered next to them. A detailed description of all parameters can be found in STIR's documentation.

@@

Enabling DOI corrections subdivides the scintillating crystal using Bresenham's line algorithm to correct for the crystal attenuation, and hence the DOI, along the LOR. If DOI correction is disabled, then the detector radius is taken to be the center of the scintillator. When ATT correction is enabled, a simple correction can be applied where the same attenuation factor is applied for the whole PSF, or a full correction can be applied where different attenuation factors are applied for each bin of the PSF~\cite{fuster_2014}. Further improvements to speed and memory can be made with image masking using the default cylinder, an attenuation map, or a mask file. The default cylinder is based on the object radius in the image volume. It is important to always set the object radius greater than or equal to the size of the object in the attenuation map or mask file when masking as the matrix weights are calculated according to this value, and failure to do so will result in an error. The projection matrix can be kept in memory or calculated per projection angle. In the latter case, the memory is released before starting calculations on a new angle, thereby reducing memory requirements but increasing computation time for iterative reconstruction algorithms.

@@

To test the pinhole-SPECT implementation in STIR, Cubresa's novel silicon-photomultiplier (SiPM)-based preclinical SPECT system --- The Spark --- was used with the single-pinhole (SPH) collimator (Cubresa Inc., Winnipeg, Canada). This system was recently characterized with the National Electrical Manufacturers Association (NEMA) NU 1-2018 Standards for Performance Measurements of Gamma Cameras, and a corresponding Geant4 Application for Tomographic Emission (GATE) Monte Carlo model was validated~\cite{strugari_nema_2022}.

@@

%The MPH collimator geometry was reverse-engineered from optical surface scans using an HP 3D structured light scanner to create a 3D computer model (HP Inc., Palo Alto, USA). The open-source CloudCompare package was used to register pinhole cones to the computer model to extract pinhole  geometry~\cite{cloudcompare}. Each row of pinholes was localized at approximately z~=~\{-15.0,~-7.5,~0.0,~7.5,~15.0\}~mm and focal spots were estimated to lie at the center of the tomographic FOV with transaxial focal spots at x~=~\{-5.0,~10.0,~0.0,~-10.0,~5.0\}~mm (note that the set ordering corresponds to the aforementioned rows). The pinholes were presumed to uniformly project over an 8$\times$8 cm$^2$ area located on the face of the SiPM-array. This provided necessary and sufficient information to simulate MPH-SPECT acquisitions using the new \texttt{GateParameterisedPinholeCollimator} class in GATE v9.2 to further test the pinhole-SPECT implementation in STIR~\cite{gate}.

@@

\begin{threeparttable}[h!]
\caption{Summary of simulated phantom acquisitions and reconstructions.\label{tab:phantom_summary}}
\footnotesize
\begin{tabular}{l l l l l l l l}
	\hline
	Subject & Activity & Acquisition & Projections & Projection matrix & Reconstruction matrix & Algorithm\footnotemark[1] & Analysis\footnotemark[2] \\ \hline

@@

To compare computation costs for different types of matrix corrections and masking, a forward projection of the IQ phantom was made with 64 views over 360~$\deg$ (see Table~\ref{tab:phantom_summary}). Memory requirements and CPU time when storing the matrix in memory were compared to calculating it per projection angle. The maximum RAM and CPU time were recorded with Ubuntu's \texttt{/usr/bin/time -v} command when calling STIR's \texttt{OSMAPOSL} program from the command line. The ordered subset expectation maximization (OSEM) algorithm was used with 8 subsets and 40 subiterations, and the matrix was calculated with no corrections (N-C), attenuation correction (ATT-C), DOI correction (DOI-C), PSF correction (PSF-C), all corrections (ATTDOIPSF-C), and all corrections with masking (ATTDOIPSFM-C) using the default cylindrical mask ($\diameter$ = 34.0 mm)~\cite{osem}.

@@

A sample sinogram of the hot rods is shown in Fig.~\ref{fig:sinogram} where Fig~\ref{fig:gate_sinogram} is the GATE result and Fig.~\ref{fig:stir_sinogram} is the forward projected STIR result including attenuation, PSF, and DOI effects. Visual agreement between these sinograms suggests that

To compare different reconstruction algorithms, the simulated IQ phantom was used to assess contrast-to-noise ratios $\mathrm{CNR}$ for each hot rod $i$:
\begin{equation}\label{eq:CNR}
	\mathrm{CNR}_i = \frac{|\mathrm{I}_i - \mathrm{I}_\mathrm{ref}|/(\mathrm{I}_i + \mathrm{I}_\mathrm{ref})}{\sigma / \mu}.
\end{equation}
Here, $\mathrm{I}_i$ is the mean intensity of the $i^\mathrm{th}$ hot rod delineated by the attenuation map, $\mathrm{I}_\mathrm{ref}$ is the mean intensity of the reference ROI central to the hot rods ($\diameter$ = 5.4 mm, $\mathrm{L}$ = 15.0 mm), and $\sigma$ and $\mu$ are the standard deviation and mean intensity, respectively, in the ROI central to the uniform volume ($\diameter$ = 18.0 mm, $\mathrm{L}$~=~11.25 mm). To elaborate, the cylindrical ROIs covered 60\% of the active diameter and 75\% of the active length based on NEMA's methodology, with the exception of the hot rod ROIs which used the entire diameter and length in analysis. Note that the coefficient of variation $\mathrm{CV}$ is represented by the denominator in Eq.~\ref{eq:CNR}:
\begin{equation}\label{eq:CV}
\mathrm{CV} = \frac{\sigma}{\mu}.
\end{equation}
The reconstruction algorithms chosen for comparison of $\mathrm{CNR}$ were OSEM, the ordered subsets one step late algorithm with median root prior (OS-OSL-MRP) using a penalization factor of $\mathrm{PF}$ = 1.0~\cite{ososlmrp}, and the ordered subsets separable paraboloidal surrogate algorithm with quadratic prior (OS-SPS-QP) using $\mathrm{PF}$ = 0.3 and relaxation parameters of $\alpha$ = 1.0 and $\gamma$ = 0.1~\cite{osspsqp}. The OS-SPS-QP algorithm was initialized with the OSEM image after 21 subiterations.

@@

An \textit{in vivo} dataset was chosen to demonstrate qualitative image quality from an investigation of novel radiotracers for Alzheimer's disease diagnosis~\cite{debay}. As summarized in Table~\ref{tab:invivo_summary}, a B6SJLF1/J mouse was administered an intravenous tail-vein injection with an iodine-123 (\ce{^{123}I})-labelled cholinesterase agent with subsequent scan commencing 2 h post-injection. The reconstructed image was visually inspected for uptake in different organs.

\begin{table}[h!]
\caption{Summary of \textit{in vivo} acquisition and reconstruction.\label{tab:invivo_summary}}
\footnotesize
\begin{tabular}{l l l l l l l l}
	\hline
	Subject & Activity & Acquisition & Projections & Projection matrix & Reconstruction matrix & Algorithm\\ \hline

@@

\begin{threeparttable}[h!]
\caption{Maximum RAM and CPU time required in SPH-SPECT OSEM reconstruction.\label{tab:computation_cost}}
\footnotesize
\begin{tabular}{l l l l l}
	\hline
	 & \multicolumn{2}{c}{Matrix in memory} & \multicolumn{2}{c}{Matrix per projection} \\ \cline{2-5}
	 Correction type\footnotemark[1] & Max RAM (MB) & CPU time (s) & Max RAM (MB) & CPU time (s) \\ \hline

@@

%-------------------------------------------------
\noindent \textbf{Figure~\ref{fig:unformity_variability} goes here.}
%-------------------------------------------------

@@

@@

\bibliographystyle{Frontiers-Harvard} %  Many Frontiers journals use the Harvard referencing system (Author-date), to find the style and resources for the journal you are submitting to: https://zendesk.frontiersin.org/hc/en-us/articles/360017860337-Frontiers-Reference-Styles-by-Journal. For Humanities and Social Sciences articles please include page numbers in the in-text citations 
%\bibliographystyle{Frontiers-Vancouver} % Many Frontiers journals use the numbered referencing system, to find the style and resources for the journal you are submitting to: https://zendesk.frontiersin.org/hc/en-us/articles/360017860337-Frontiers-Reference-Styles-by-Journal
\bibliography{test}

@@

%-------------------------------------------------
\begin{figure}[h!]
\begin{center}
\includegraphics[height=6cm]{Figures/STIR-UsersGuide_PinholeSPECTUB}
\end{center}
\caption{\texttt{PinholeSPECTUB} system of reference and sign criteria illustrated for a polygonal collimator setup. Note that the projection matrix adheres to STIR's coordinate system as indicated by the x, y, and z axes. The detector and collimator use a rotating frame of reference where the transaxial x' and axial z' axes coincide with STIR's axes when the detector is at 0 $\mathrm{\deg}$. The collimator uses a right-handed coordinate system as indicated by the y' axis which points toward the detector. Further information is given in the text and in STIR's documentation.}
\label{fig:PinholeSPECTUB_coords}
\end{figure}

%-------------------------------------------------
\begin{figure}[h!]
\begin{center}
\includegraphics[height=6cm]{Figures/HotRod_sinograms_y130}
\end{center}
\caption{Projection of IQ phantom hot rod region displayed in a 2D sinogram arrangement showing the GATE simulated data (\textbf{left}) and the STIR forward projection of the radioactive source distribution adding PSF, DOI, and ATT degradations (\textbf{right}).}
\label{fig:PinholeSPECTUB_coords}
\end{figure}

%-------------------------------------------------
\begin{figure}[h!]
\begin{center}
\includegraphics[width=\textwidth]{Figures/HotRod_CNR}
\end{center}
\caption{IQ phantom hot rod CNR plots comparing OSEM (solid line), OS-OSL with median root prior (dotted line), and OS-SPS with quadratic prior (dashed line) over subiterations for the 1 mm hot rod (\textbf{left}), 3 mm hot rod (\textbf{middle}), and 5 mm hot rod (\textbf{right}). All images were reconstructed with no matrix corrections and seven subsets, and the OS-SPS-QP reconstruction was initialized using the OSEM image after 21 subiterations. Hot rod contrast was calculated relative to the central inter-rod region void of \ce{^{99m}Tc}, and CV was calculated in the uniform \ce{^{99m}Tc} region.}
\label{fig:CNR}
\end{figure}

%-------------------------------------------------
\begin{figure}[h!]
\begin{center}
\includegraphics[width=\textwidth]{Figures/OSEM_reconstructions}
\end{center}
\caption{Simulated data. Axial sum of normalized OSEM images after 35 subiterations with no matrix corrections and seven subsets for the IQ phantom hot rods (\textbf{left}), mouse-sized NEMA line source phantom (\textbf{middle}), and volumetric cylinder (\textbf{right}). The IQ phantom image was summed over the length of the hot rods whereas the other images were summed over the entire length of the reconstructed image. Note the expected distributions of \ce{^{99m}Tc}.}
\label{fig:reconstructions}
\end{figure}

%-------------------------------------------------
\begin{figure}[h!]
\begin{center}
\includegraphics[width=0.5\linewidth]{Figures/Resolution.pdf}
\end{center}
\caption{SPECT spatial resolution with scatter in the mouse-sized NEMA triple line source phantom using the OSEM algorithm with seven subsets.}\label{fig:resolution}
\end{figure}

%-------------------------------------------------
\begin{figure}
    \centering
    \begin{minipage}[t]{0.5\textwidth}
        \includegraphics[width=\linewidth]{Figures/VSAC_Uniformity.pdf}
    \end{minipage}%
    \begin{minipage}[t]{0.5\textwidth}
        \includegraphics[width=\linewidth]{Figures/VSAC_CV.pdf}
    \end{minipage}
\caption{SPECT uniformity (\textbf{left}) and variability (\textbf{right}) in the volumetric cylinder using the OSEM algorithm with 7 subsets.}\label{fig:uniformity_variability}
\end{figure}

%-------------------------------------------------
\begin{figure}[h!]
\begin{center}
%\includegraphics[width=0.5\linewidth]{Figures/InVivo.pdf}
\end{center}
\caption{Sagittal fused SPECT/CT of the \textit{in vivo} mouse with a window ranging from 0 to 1. The \ce{^{123}I} distribution shows a nonpersistent tracer in the brain.}\label{fig:invivo}
\end{figure}

%-------------------------------------------------
\begin{figure}[h!]
\begin{center}
\includegraphics[width=\textwidth]{Figures/DOI_CylVol_bug}
\end{center}
\caption{Simulated data. Slices of the volumetric cylinder thresholded between 0 and 1.}
\label{fig:DOIbug}
\end{figure}

%%% If you don't add the figures in the LaTeX files, please upload them when submitting the article.
%%% Frontiers will add the figures at the end of the provisional pdf automatically
%%% The use of LaTeX coding to draw Diagrams/Figures/Structures should be avoided. They should be external callouts including graphics.

>>>>

\documentclass[utf8]{FrontiersinVancouver}

\usepackage{url,hyperref,lineno,microtype}
\usepackage[onehalfspacing]{setspace}
\usepackage[version=4]{mhchem} % for element symbols
\usepackage{wasysym} % for diameter symbol
\usepackage{multirow}
\usepackage{threeparttable}
%\usepackage{subcaption}	% will use only figures with built-in subfigures for automatic numbering of figure order

% Raise version of output document to 1.7 ti suppress warning
% PDF inclusion: found PDF version <1.7> but at most version <1.5> allowed
\pdfminorversion=7

@@

\def\keyFont{\fontsize{8}{11}\helveticabold}
\def\firstAuthorLast{Strugari {et~al.}} %use et al only if is more than 1 author

@@

\def\Address{
$^{1}$Biomedical MRI Research Laboratory, IWK Health Centre, Halifax, Canada\\
$^{2}$Department of Physics \&  Atmospheric Science, Dalhousie University, Halifax, Canada\\
$^{3}$Neuroimaging Group, Barcelona$\mathrm{\beta}$eta Brain Research Center, Barcelona, Spain\\
$^{4}$Institute of Nuclear Medicine, University College London, London, UK\\
$^{5}$Department of Diagnostic Radiology, Dalhousie University, Halifax, Canada
}
% The Corresponding Author should be marked with an asterisk
% Provide the exact contact address (this time including street name and city zip code) and email of the corresponding author
\def\corrAuthor{Matthew Strugari}
\def\corrEmail{matthew.strugari@dal.ca}

@@

The design of the collimator in terms of hole-size, material, and overall geometry, among other factors, affects the spatial resolution and sensitivity of a SPECT system. A number of designs exist including but not limited to parallel-hole, fan-beam, converging and diverging, slit-slat, coded-aperture, and single- and multi-pinhole collimators. Therefore, the choice of collimator design is application dependent in order to channel photons of different energies, magnify or minify images, or select between imaging quality and imaging speed~\citep{van_mullekom_group_collimators_2021}. Although parallel-hole and fan-beam collimators are conventionally used when imaging small fields-of-view (FOVs), pinhole collimators can provide a superior resolution-sensitivity tradeoff~\citep{islamian_advances_2015}. Besides the successful application of pinhole-SPECT systems in small-animal imaging, there has been a resurgence in the use of pinhole collimators for clinical cardiac and brain studies and when imaging small FOVs~\citep{ozsahin_clinical_2020}.

While the use of pinhole-SPECT has regained popularity in clinical and preclinical investigations of molecular imaging agents, there are no open-source software solutions available for reconstructing pinhole-SPECT datasets.  However, recent efforts have led to the integration of a 3D SPECT system matrix modelling library for pinhole collimators into the open-source Software for Tomographic Image Reconstruction (STIR). The STIR package is an object-oriented library implemented in C++ that provides a framework for research in the processing and reconstruction of emission tomography studies~\citep{thielemans_stir_2012}. Initially written for support of positron emission tomography (PET) data, STIR was previously extended to handle SPECT data with parallel- and converging-hole collimators~\citep{marti_fuster_integration_2013, marti_fuster_evaluation_2013}. The expansion of STIR's support for pinhole collimators marks the first open-source platform for reconstructing pinhole-SPECT datasets which is important in the advancement of molecular imaging techniques and technologies.

@@

This work aims to demonstrate the forthcoming capabilities of STIR's support for pinhole-SPECT datasets. This was achieved by integrating parts of the SPECT Reconstruction Library developed at the University of Barcelona (SRL-UB) into STIR~\citep{falcon_metodos_1999, cot_absolute_2005, pareto_geometrical_2002, pareto_iterative_2003}. The library enables corrections for the spatially variant collimator-detector response and attenuation by incorporating their effects into the system matrix.

@@

Similar to the original \texttt{SPECTUB} implementation, the new pinhole-SPECT implementation is referred to as \texttt{PinholeSPECTUB} and includes a dedicated reader for SPECT projection data in interfile format~\citep{todd-pokropek_file_1992}. The pinhole-SPECT interfile reader utilizes the projection matrix size, pixel scaling factor, and detector radius according to the face of the scintillating crystal. Calculation of the system matrix is executed with the \texttt{ProjMatrixByBinPinholeSPECTUB} projector class derived from the existing STIR \texttt{ProjMatrixByBin} class which utilizes a detector and collimator text file in addition to the usual STIR parameter file. The parameter file is a text file which uses an Interfile-like syntax. It is composed of keywords corresponding to the names of the various reconstruction and matrix parameters with the values entered next to them. A detailed description of all parameters can be found in STIR's documentation.

@@

Enabling DOI corrections subdivides the scintillating crystal using Bresenham's line algorithm to calculate the crystal attenuation and DOI along the LOR. If DOI correction is disabled, then half the crystal thickness is added to the detector radius.  When ATT correction is enabled, a simple correction can be applied where the same attenuation factor is applied for the whole PSF, or a full correction can be applied where different attenuation factors are applied for each bin of the PSF~\citep{marti_fuster_evaluation_2013}. Further improvements to speed and memory can be made with image masking using the default cylinder, an attenuation map, or a mask file. The default cylinder is based on the object radius in the image volume. It is important to always set the object radius greater than or equal to the size of the object in the attenuation map or mask file when masking as the matrix weights are calculated according to this value, and failure to do so will result in an error. The projection matrix can be kept in memory or calculated per projection angle. In the latter case, the memory is released before starting calculations on a new angle, thereby reducing memory requirements but increasing computation time for iterative reconstruction algorithms.

@@

To test the pinhole-SPECT implementation in STIR, Cubresa's novel silicon-photomultiplier (SiPM)-based preclinical SPECT system --- The Spark --- was used with a single-pinhole (SPH) collimator (Cubresa Inc., Winnipeg, Canada). This system was recently characterized with the National Electrical Manufacturers Association (NEMA) NU 1-2018 Standards for Performance Measurements of Gamma Cameras, and a corresponding Geant4 Application for Tomographic Emission (GATE) Monte Carlo model was validated~\citep{strugari_nema_2022}.

@@

%The MPH collimator geometry was reverse-engineered from optical surface scans using an HP 3D structured light scanner to create a 3D computer model (HP Inc., Palo Alto, USA). The open-source CloudCompare package was used to register pinhole cones to the computer model to extract pinhole  geometry~\citep{cloudcompare}. Each row of pinholes was localized at approximately z~=~\{-15.0,~-7.5,~0.0,~7.5,~15.0\}~mm and focal spots were estimated to lie at the center of the tomographic FOV with transaxial focal spots at x~=~\{-5.0,~10.0,~0.0,~-10.0,~5.0\}~mm (note that the set ordering corresponds to the aforementioned rows). The pinholes were presumed to uniformly project over an 8$\times$8 cm$^2$ area located on the face of the SiPM-array. This provided necessary and sufficient information to simulate MPH-SPECT acquisitions using the new \texttt{GateParameterisedPinholeCollimator} class in GATE v9.2 to further test the pinhole-SPECT implementation in STIR~\citep{gate}.

@@

\begin{threeparttable}[ht!]
\caption{Summary of simulated phantom acquisitions and reconstructions.\label{tab:phantom_summary}}
\footnotesize
\begin{tabular}{l l l l l l l l}
	\hline
	Subject & Activity & Acquisition & Projections & Projection matrix & Reconstruction matrix & Algorithm\footnotemark[1] & Analysis\footnotemark[2] \\ \hline

@@

To compare computation costs for different types of matrix corrections and masking, a forward projection of the IQ phantom was made with 64 views over 360~$\deg$ (see Table~\ref{tab:phantom_summary}). Memory requirements and CPU time when storing the matrix in memory were compared to calculating it per projection angle. The maximum RAM and CPU time were recorded with Ubuntu's \texttt{/usr/bin/time -v} command when calling STIR's \texttt{OSMAPOSL} program from the command line. The ordered subset expectation maximization (OSEM) algorithm was used with 8 subsets and 40 subiterations, and the matrix was calculated with no corrections (N-C), attenuation correction (ATT-C), DOI correction (DOI-C), PSF correction (PSF-C), all corrections (ATTDOIPSF-C), and all corrections with masking (ATTDOIPSFM-C) using the default cylindrical mask ($\diameter$ = 34.0 mm)~\citep{hudson_accelerated_1994}.

@@

Sample sinograms of the IQ phantom hot rods are shown in Fig.~\ref{fig:HotRod_sinos} from the GATE simulation and the STIR forward projection including ATT, DOI, and PSF effects. Visual agreement between these sinograms supports that the implementation of the \texttt{PinholeSPECTUB} projector matrix in STIR is suitable for pinhole-SPECT datasets.

To compare different reconstruction algorithms, the simulated IQ phantom was used to assess contrast-to-noise ratios $\mathrm{CNR}$ for each hot rod $i$:
\begin{equation}\label{eq:CNR}
	\mathrm{CNR}_i = \frac{|\mathrm{I}_i - \mathrm{I}_\mathrm{ref}|/(\mathrm{I}_i + \mathrm{I}_\mathrm{ref})}{\sigma / \mu}.
\end{equation}
Here, $\mathrm{I}_i$ is the mean intensity of the $i^\mathrm{th}$ hot rod delineated by the attenuation map, $\mathrm{I}_\mathrm{ref}$ is the mean intensity of the reference ROI central to the hot rods ($\diameter$ = 5.4 mm, $\mathrm{L}$ = 15.0 mm), and $\sigma$ and $\mu$ are the standard deviation and mean intensity, respectively, in the ROI central to the uniform volume ($\diameter$ = 18.0 mm, $\mathrm{L}$~=~11.25 mm). To elaborate, the cylindrical ROIs covered 60\% of the active diameter and 75\% of the active length based on NEMA's methodology, with the exception of the hot rod ROIs which used the entire diameter and length in analysis. Note that the coefficient of variation $\mathrm{CV}$ is represented by the denominator in Eq.~\ref{eq:CNR}:
\begin{equation}\label{eq:CV}
\mathrm{CV} = \frac{\sigma}{\mu}.
\end{equation}
The reconstruction algorithms chosen for comparison of $\mathrm{CNR}$ were OSEM, the ordered subsets one step late algorithm with median root prior (OS-OSL-MRP) using a penalization factor of $\mathrm{PF}$ = 1.0~\citep{green_use_1990}, and the ordered subsets separable paraboloidal surrogate algorithm with quadratic prior (OS-SPS-QP) using $\mathrm{PF}$ = 0.3 and relaxation parameters of $\alpha$ = 1.0 and $\gamma$ = 0.1~\citep{ahn_globally_2003}. The OS-SPS-QP algorithm was initialized with the OSEM image after 21 subiterations.

@@

An \textit{in vivo} dataset was chosen to demonstrate qualitative image quality from an investigation of novel radiotracers for Alzheimer's disease diagnosis~\citep{macdonald_early_2014, debay_targeting_2017}. As summarized in Table~\ref{tab:invivo_summary}, a B6SJLF1/J mouse was administered an intravenous tail-vein injection with an iodine-123 (\ce{^{123}I})-labelled cholinesterase agent with subsequent scan commencing 2 h post-injection. The reconstructed image was visually inspected for uptake in different organs.

\begin{table}[ht!]
\caption{Summary of \textit{in vivo} acquisition and reconstruction.\label{tab:invivo_summary}}
\footnotesize
\begin{tabular}{l l l l l l l l}
	\hline
	Subject & Activity & Acquisition & Projections & Projection matrix & Reconstruction matrix & Algorithm\\ \hline

@@

\begin{threeparttable}[ht!]
\caption{Maximum RAM and CPU time required in SPH-SPECT OSEM reconstruction.\label{tab:computation_cost}}
\footnotesize
\begin{tabular}{l l l l l}
	\hline
	 & \multicolumn{2}{c}{Matrix in memory} & \multicolumn{2}{c}{Matrix per projection} \\ \cline{2-5}
	 Correction type\footnotemark[1] & Max RAM (MB) & CPU time (s) & Max RAM (MB) & CPU time (s) \\ \hline

@@

%-------------------------------------------------
\noindent \textbf{Figure~\ref{fig:uniformity_variability} goes here.}
%-------------------------------------------------

@@

%-------------------------------------------------
\noindent \textbf{Figure~\ref{fig:invivo} goes here.}
%-------------------------------------------------

@@

%\bibliographystyle{Frontiers-Harvard} %  Many Frontiers journals use the Harvard referencing system (Author-date), to find the style and resources for the journal you are submitting to: https://zendesk.frontiersin.org/hc/en-us/articles/360017860337-Frontiers-Reference-Styles-by-Journal. For Humanities and Social Sciences articles please include page numbers in the in-text citations 
\bibliographystyle{vancouver} % Many Frontiers journals use the numbered referencing system, to find the style and resources for the journal you are submitting to: https://zendesk.frontiersin.org/hc/en-us/articles/360017860337-Frontiers-Reference-Styles-by-Journal
\bibliography{references}

@@

%-------------------------------------------------
\begin{figure}[ht!]
\begin{center}
\includegraphics[height=6cm]{Figures/STIR-UsersGuide_PinholeSPECTUB}
\end{center}
\caption{\texttt{PinholeSPECTUB} system of reference and sign criteria illustrated for a polygonal collimator setup. Note that the projection matrix adheres to STIR's coordinate system as indicated by the x, y, and z axes. The detector and collimator use a rotating frame of reference where the transaxial x' and axial z' axes coincide with STIR's axes when the detector is at 0 $\mathrm{\deg}$. The collimator uses a right-handed coordinate system as indicated by the y' axis which points toward the detector. Further information is given in the text and in STIR's documentation.}
\label{fig:PinholeSPECTUB_coords}
\end{figure}

%-------------------------------------------------
\begin{figure}[ht!]
\begin{center}
\includegraphics[width=0.71\textwidth]{Figures/HotRod_sinograms_y130}
\end{center}
\caption{Projection of IQ phantom hot rod region displayed in a 2D sinogram arrangement showing the GATE simulated data (\textbf{left}) and the STIR forward projection of the radioactive source distribution adding PSF, DOI, and ATT degradations (\textbf{right}).}
\label{fig:HotRod_sinos}
\end{figure}

%-------------------------------------------------
\begin{figure}[ht!]
	\begin{center}
		\includegraphics[width=\textwidth]{Figures/OSEM_reconstructions}
	\end{center}
	\caption{Simulated data. Axial sum of normalized OSEM images after 35 subiterations with no matrix corrections and seven subsets for the IQ phantom hot rods (\textbf{left}), mouse-sized NEMA line source phantom (\textbf{middle}), and volumetric cylinder (\textbf{right}). The IQ phantom image was summed over the length of the hot rods whereas the other images were summed over the entire length of the reconstructed image. Note the expected distributions of \ce{^{99m}Tc}.}
	\label{fig:reconstructions}
\end{figure}

%-------------------------------------------------
\begin{figure}[ht!]
\begin{center}
\includegraphics[width=\textwidth]{Figures/HotRod_CNR}
\end{center}
\caption{IQ phantom hot rod CNR plots comparing OSEM (solid line), OS-OSL with median root prior (dotted line), and OS-SPS with quadratic prior (dashed line) over subiterations for the 1 mm hot rod (\textbf{left}), 3 mm hot rod (\textbf{middle}), and 5 mm hot rod (\textbf{right}). All images were reconstructed with no matrix corrections and seven subsets, and the OS-SPS-QP reconstruction was initialized using the OSEM image after 21 subiterations. Hot rod contrast was calculated relative to the central inter-rod region void of \ce{^{99m}Tc}, and CV was calculated in the uniform \ce{^{99m}Tc} region.}
\label{fig:CNR}
\end{figure}

%-------------------------------------------------
\begin{figure}[ht!]
\begin{center}
\includegraphics[width=0.49\linewidth]{Figures/Resolution}
\end{center}
\caption{SPECT spatial resolution with scatter in the mouse-sized NEMA triple line source phantom using the OSEM algorithm with seven subsets.}\label{fig:resolution}
\end{figure}

%-------------------------------------------------
\begin{figure}[ht!]
\begin{center}
\includegraphics[width=\linewidth]{Figures/Uniformity_Variability}
\end{center}
\caption{SPECT uniformity (\textbf{left}) and variability (\textbf{right}) in the volumetric cylinder using the OSEM algorithm with 7 subsets.}\label{fig:uniformity_variability}
\end{figure}

%-------------------------------------------------
%\begin{figure}
%    \centering
%    \begin{minipage}[t]{0.5\textwidth}
%        \includegraphics[width=\linewidth]{Figures/VSAC_Uniformity.pdf}
%    \end{minipage}%
%    \begin{minipage}[t]{0.5\textwidth}
%        \includegraphics[width=\linewidth]{Figures/VSAC_CV.pdf}
%    \end{minipage}
%\caption{SPECT uniformity (\textbf{left}) and variability (\textbf{right}) in the volumetric cylinder using the OSEM algorithm with 7 subsets.}\label{fig:uniformity_variability}
%\end{figure}

%-------------------------------------------------
\begin{figure}[ht!]
\begin{center}
%\includegraphics[width=0.5\linewidth]{Figures/InVivo.pdf}
\end{center}
\caption{Sagittal fused SPECT/CT of the \textit{in vivo} mouse with a window ranging from 0 to 1. The \ce{^{123}I} distribution shows a nonpersistent tracer in the brain.}\label{fig:invivo}
\end{figure}

%-------------------------------------------------
\begin{figure}[ht!]
\begin{center}
\includegraphics[width=\textwidth]{Figures/DOI_CylVol_bug}
\end{center}
\caption{Slices of the volumetric cylinder reconstructed with OSEM in 35 subiterations using DOI correction where image values were thresholded between 0 and 1. The DOI correction bug is illustrated in the central axial (\textbf{left}), coronal (\textbf{middle}), and sagittal (\textbf{right}) planes. The bug affects voxels within a small angle from the pinhole axis as seen along the pinhole trajectory in a 270 $\deg$ counter-clockwise acquisition starting at 180 $\deg$. The axial view shows the formation of a multi-armed cross or `star shot' artifact, and all views show the compounding effect at the isocenter due to the intersection of LORs affected by the bug.}
\label{fig:DOIbug}
\end{figure}
